Variance reduction for MCMC methods via martingale representations

### 2020-04-11 aap/Revision/vr-mcmc-revision-new-new.tex
update

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

@@

Y_p, & \texttt{ with probability } \alpha_p(x,Y_p),

@@

x, &  \texttt{ with probability } 1-\alpha_p(x,Y_p),

@@

\alpha_p(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.

@@

%On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)

@@

&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l}).

@@

for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator

@@

The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way to estimate them is to use nonparametric regression.

@@

for $k=1,\ldots,K,$ and some $l_0$ with $l_0+q\leq N+n,$

@@

\widehat{A}_{q,k}(y):=

@@

\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].

@@

\label{eq:PVarA}

@@

\bar A_{q,k}:=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2

@@

 as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover, it follows from \eqref{eq:PVarA} that

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n$ and $k=1,\ldots,K$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$\section{Analysis of variance reduced ULA}

@@

H_k(x):=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},

@@

\mathbf{H}_\mathbf{k}(\mathbf{x}):=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.

@@

M_{K,n}^N(f) & :=\frac{1}{n}\sum_{ 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})

@@

A_{q,\mathbf{k}}(y):=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).

@@

\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz continuous, that is, there exists $L < \infty$ such that

@@

\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
&=& 
\label{eq:a-pol}
\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),

@@

\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].

@@

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
\end{eqnarray}
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$ to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$. Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and depict boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$. Next we compute the cost-variance ratios
\[
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
\]
where variances in the above definition are computed using $24$ independent trajectories of the length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$. In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters. In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of a linear functional $f$. In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of a quadratic functional $f.$
Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for a mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$ 
 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig3}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig4}}
\caption{ Cost-variance ratios as a function of the truncation level for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig5}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig6}}
\caption{ Cost-variance ratios as a function of the truncation level for  a mixture of two-dimensional  Gaussian distribution and different test functions.
\label{fig:GMM-costvar}}
\end{figure}

\subsection{Banana shape distribution}
The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
\[
\varphi(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
\]
where $p > 0$ and $b > 0$ are parameters; $b$ accounts for the curvature of density’s level sets. The potential $U$ is given by
\[
U(x_1, . . . , x_d) = \frac{x_1^2}{2} + (x_2 + bx_1^2 - pb)^2 + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
\]
The quantity of interest is the expectation of $f(x) = x_2$. We set $p = 100, b = 0.1$ and consider $d=8$. We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$. We use $T=10$ independent training trajectories, each of size $n=10^4$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.

>>>>

Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian Statistics. Suppose that we aim at computing $\pi(f) = \pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$. Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution. Given an independent identically distributed observations $X_1,\ldots,X_n$ form $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$. The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand. The first way to obtain tighter estimate $\hat{\pi}(f)$ is simply to increase the sample size $n$. Unfortunately, this solution might be prohibitively costly, especially when the dimension $d$ is large enough or sampling from $\pi$ is complicated. An alternative approach is to decrease $\sigma^2(f)$ by constructing a new Monte-Carlo experiment with the same expectation as the original one, but with a lower variance. Methods to achieve this are known as variance
reduction techniques. Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par

One of the popular approaches to variance reduction is the so called control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) - \zeta$ is small. The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

@@

Y_p, & \texttt{ with probability } \alpha(x,Y_p),

@@

x, &  \texttt{ with probability } 1-\alpha(x,Y_p),

@@

\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.

@@

On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)

@@

&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].

@@

for any integrable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator

@@

The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.

@@

for some $l_0$ with $l_0+q\leq N+n,$

@@

\widehat{A}_{q,k}(y)=

@@

\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].

@@

@@

\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2

@@

 as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover,

@@

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$. At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-variances is of order
\label{eq:cost_theoretical}

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
\section{Analysis of variance reduced ULA}

@@

H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},

@@

\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.

@@

M_{K,n}^N(f) & =\sum_{ 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})

@@

A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).

@@

\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that

@@

\label{eq:a_hat}
\widehat a_{r,\mathbf{k}} (x) = \CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),

@@

\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].

@@

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-\frac{\|x-\mu\|^2}{2}} + \rme^{-\frac{\|x+\mu\|^2}{2}}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the quantity
\begin{eqnarray*}
\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}
\end{eqnarray*}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. For different $n_{\mathrm{trunc}}$ and $K$ we estimate $\mathrm{cost}(\widehat \pi_{K,n}^N)$ as 
 \[
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times d^{K}
 \]
 since the representations of $a_{r,\mathbf{k}} (x)$ might be pre-computed using formulas \eqref{eq:a_hat}. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig2}}
\caption{VR cost of MDCV algorithm compared to vanilla ULA on Gaussian potential
\label{fig:gaus_cost}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig2}}
\caption{VR cost of MDCV algorithm compared to vanilla ULA on Gaussian mixture potential
\label{fig:gmm_cost}}
\end{figure}

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
\]
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

\subsection{Banana shape distribution}
The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
\[
\varphi(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
\]
where $p > 0$ and $b > 0$ are parameters; $b$ accounts for the curvature of density’s level sets. The potential $U$ is given by
\[
U(x_1,\ldots, x_d) = \frac{x_1^2}{2} + \frac{(x_2 + bx_1^2 - pb)^2}{2} + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
\]
The quantity of interest is the expectation of $f(x) = x_2$. We set $p = 100, b = 0.1$ and consider $d=8$. We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$. We use $T= 5 \times 10^4$ independent training trajectories, each of size $n=50$ with the burn-in $n_{\text{burn}} = 100$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
